Reformulating phylogenetic mixed models to improve flexibility and speed

### 2022-08-03 response.md

As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).

**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast existing models in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application. We have provided an example application to empirical data in a package vignette.**

The benchmarking of the new method is interesting but is mixing too many confounding factors linked with the different implementations of the various packages (see also comment from reviewer 2). What is the performance of the new model all else being equal? How much of the performance results is actually due to differences the optimization among the different R packages?

**For models that are sufficiently flexible to fit a model structure of given complexity (e.g. brms/pez), all of the models are mathematically equivalent; the difference is in the precise formulation and implementation, leading to the observed significant differences in speed. The most direct comparison is between `phyloglmm` and the `phyr` or `pez` packages: all three methods are built on the mixed-models machinery of `lme4`, but `phyloglmm` modifies the Z matrix while `phyr` and `pez` modify the correlation structure of the random effects**.

@@

**Thank you.**

However, I think the article could fail to reach the extent of its audience by missing several key elements:

- Because that work isn’t a theoretical work but instead the description of an applied statistics method, adding a few examples, such as the one provided at l68-73, would help the reader understanding the value of those models. I think examples should particularly be aimed at the “Simulation” § l132-155, where they could shed light on the subtle differences between the parameters

**We have added an example in the vignette for this package.**

@@

- It would also be beneficial to add an application of the method with biological data directly in the article. The more your article is associated with biological ideas, the more users you will reach in my opinion.

**Thanks you for the suggestion. This paper is written to be an application paper rather than a research paper. We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository (we could add this if the editor thought it was appropriate, but we're up against the length limits for an application paper and would prefer not to open the can of worms of making this into a research paper ...**

- Naming that class of models (even though it is still pglmm) could help the reader follow the results section and identify the method as well.

**We have now renamed everything under phylogenetic regression**

@@

These remarks might sound cosmetic but I think they might help to reach a broader audience and help all kind of users getting started with those models.

I recommend the authors to think on those ideas to improve the visibility of their method and the journal to ask them to do so in order to increase the impact of the article.

In a purely scientific perspective, I have no objections to publish that article.

**These are very useful suggestions. We have included a tutorial example in the released version on zenodo/github; the link is included n the paper.**

@@

1. What is the computational complexity of the proposed method? Does the alternative formulation help avoid matrix inversion such that the complexity is linear with respect to the number of species? Note that this is the case for many phylogenetic packages including phylolm and MCMCglmm. I am aware that it is hard or even impossible to achieve linear-complexity for all models. However, could linear-complexity be achieved by the proposed methods for some simple models such as the standard phylogenetic linear regression (equation 1)?

**Exactly, we are avoiding the matrix inversion step.**

2. In the paper, the authors replace the package MCMCglmm by the package brms. Although the implementation of MCMCglmm is not as efficient as brms, the complexity of MCMCglmm is linear with respect to the number of species while brms should be cubic (since it doesn't have a special method to avoid matrix inversion). It may be a good idea to add a simulation with a series of trees from dozens to thousands of species for showing the scalability of these methods (including MCMCglmm).

**We have now added MCMCglmm to our set of simulation. For our simple simulation test, MCMCglmm was generally faster than brms, although both Bayesian approaches (MCMCglmm/brms) were still significantly slower than frequentist methods, and scaled slightly worse.**

@@

3. The models in simulations are not sufficiently described. What are the formulae for the phylogenetic random intercept variance, phylogenetic random slope variance, and covariance between phylogenetic random intercept and slope, and so on? Do they depend on the phylogenetic tree? The authors provide a clear explanation for how their proposed method can be applied to the standard phylogenetic regression (equation 1) but do not include it in the simulation section. I suggest the authors add this model to the simulation part for illustrating the applicability of their proposed method for this simple case.

**FIXME: make sure we state that the simulations follow the estimation model exactly, so that all the formulas for simulation are the same as those for estimation (and maybe emphasize that the Z/Sigma setup generalizes standardized phylogenetic regression to all of the more complex cases).**

### 2022-08-08 phyloglmm_ms_resubmit.tex
fixing up plots (not quite done)
pushing minor edits in ms and updated todo

Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$))
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
    Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sstime.pdf}
  \caption{Comparison of single-group model computational speed.}
\label{ssplot_speed}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
  \caption{Comparison of coverage probability for fixed effect parameters. Models matching the simulation model (\pkg{\plmefour}, \pkg{\pgTMB} and \pkg{brms}) have coverage near the nominal value of 0.95. The black line shows the nominal coverage; the blue ribbon shows the 95\% binomial confidence interval around the nominal coverage based on fits to 100 simulated data sets.
%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
  }
\label{ssplot_coverage}
\end{figure}
\end{center}

>>>>

Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)).
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
    Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/csstime.pdf}
  \caption{Comparison of single-group model computational speed.}
\label{ssplot_speed}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
  \caption{Comparison of coverage probability for fixed effect parameters. Models matching the simulation model (\pkg{\plmefour}, \pkg{\pgTMB} and \pkg{brms}) have coverage near the nominal value of 0.95. The black line shows the nominal coverage; the blue ribbon shows the 95\% binomial confidence interval around the nominal coverage based on fits to 100 simulated data sets.
%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
  }
\label{ssplot_coverage}
\end{figure}
\end{center}
